---
title: "Intro to R - Relational Data"
author: "A.Kirkham"
date: "10/02/2020"
#output: html_document
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This guide to relational data has been built following slides from the NHS-R community at:  
https://github.com/nhs-r-community/intro_r/blob/master/07-workshopv2_joins.pdf

The below code and output are generated from following these slides.

All of the resources used here are found on the NHS-R Community Github pages:
https://github.com/nhs-r-community/intro_r


**Now follows a workthrough of the slides.**

## Relational Data
It is rare to find all of the data you require for an analysis within a single table.

Typically you will have to link two (or more) tables together by matching on common "key" variables.  
- Joins in SQL.

The main focus here will be on left (outer) joins.  
The syntax is similar for other types of joins.

## Left Join
```{r, eval=FALSE}
table_1 %>%
  left_join(table_2, by = "x")
```

Where we:  
1: Keep the structure of table_1  
2: ... and match to observations in table_2  
3: With a "key" variable (common to both tables).

We're going to join two tables - one with cases of tuberculosis by country, one with population by country.  
From this new table we can derive a rate.

Import: `tb_cases.csv`, and `tb_pop.csv`.

```{r, message=FALSE}
library(tidyverse)
library(readr)
tb_cases <- read_csv("tb_cases.csv")
tb_popul <- read_csv("tb_pop.csv")
```

Then we can do the following, whilst ensuring that we:  
1: Keep the structure of the `tb_cases` data frame;  
2: Match to rows in `tb_popul`;  
3: Based on "country"

```{r}
tb_cases %>%
  left_join(tb_popul, by = "country")
```

**But, oh no! Duplicates!**  
For every value of Brazil in tb_cases there are four in `tb_popul`...

### Join on multiple rows
Using the vector syntax that we discussed in the Addendum of Intro_to_R_dplyr we can combine multiple matching variables for the join.

```{r}
tb_cases %>%
  left_join(tb_popul, by = c("country", "year"))
```

### Joining with different names
If two tables have **different names** for the **same variable**:

```{r, eval=FALSE}
tb_cases %>%
  left_join(bad_names, by = c("country" = "Place", "Year" = "yr"))
# This won't run of course - it is fictional data. The left side of the = is the originating table (cases), whereas the right side of the = relates to the joining table (bad_names).
```

## Other dplyr joins
Some other dplyr joins include:

* dplyr::right_join(a,b,by="x1")
  + Join matching rows from a to b.
* dplyr::inner_join(a,b,by="x1")
  + Join data. Retain only rows in both sets.
* dplyr::full_join(a,b,by="x1")
  + Join data. Retain all values, all rows.