---
title: "Intro to R - naming objects, and assignments"
author: "A.Kirkham"
date: "10/02/2020"
#output: html_document
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
This is a continuation / supplement to the previous guide to wrangling data uing the dplyr() [dee-ply-r] package (and also ggplot).

This guide to naming objects has been built following slides from the NHS-R community at:  
https://github.com/nhs-r-community/intro_r/blob/master/06-workshopv2_objects.pdf

The below code and output are generated from following these slides.

All of the resources used here are found on the NHS-R Community Github pages:
https://github.com/nhs-r-community/intro_r


**Now follows a workthrough of the slides.**

Note that this code will not work without the below being run first (if you have cleared down the environment):

```{r, message=FALSE}
library(tidyverse)
library(readr)
beds_data <- read_csv("beds_data.csv", col_types = cols(date = col_date(format = "%d/%m/%Y")), 
                      skip = 3)
beds_data
```
 
Revisiting the dplyr slides we produced the following code:
```{r}
beds_data %>%
  group_by(date) %>%
  summarise(
    mean_beds = mean(beds_av, na.rm = T)
  ) %>%
  ungroup()
```

All fine and good, but something to consider is that the **output = object = new data frame**.

## Next step: Visualise
Often we want to actually visualise the output, not just look at it in a block of values in R.

### Piece it together
```{r, eval=FALSE}
beds_data %>%
  group_by(date) %>%
  summarise(
    mean_beds = mean(beds_av, na.rm = T)
  ) %>%
  ungroup()
```

Then... take this new output/object/data frame into:
`ggplot(data = [this new data])`

So, we could do something like:

```{r, eval=FALSE}
ggplot(data = beds_data %>%
         group_by(date)%>%
         summarise(
           mean_beds = mean(beds_av, na.rm = T))
       ) +
  geom...
```

**But it is often better to keep wrangling separate.**.

### And it is always better to:
**Keep your code** as **readable** as possible.

**The solution is to name your code elements.**

We can take all of the code:  
```{r, eval=FALSE}
beds_data %>%
  group_by(date) %>%
  summarise(
    mean_beds = mean(beds_av, na.rm = T)
  ) %>%
  ungroup()
```

and assign a **name** to it. 

### Good object names
Naming something is tricky. I always tend to name things "Gerald", I don't know why (AK).  
However, good object names are:

* Descriptive;
* Short(ish);
* Consistent with other names.

So something like `beds_ts` may work well in this instance.

Although this is being done within tidyverse assignment of code to a name/variable is the same as in all R scripting "<-".  
*Alternatively use the key combination of alt-.*

This is known as the assignment operator.

As such, we can associate all of the previous code to a single name. This means that we can repeatedly call it, we don't have to repeatedly type it out, and most importantly we keep all wrangling separate and consistent.

```{r}
beds_ts <- beds_data %>%
            group_by(date) %>%
            summarise(
              mean_beds = mean(beds_av, na.rm = T)
            ) %>%
            ungroup()
```

If we now call `beds_ts` we get the output expected:
```{r}
beds_ts
```

You will find that `beds_ts` is now also in the Environment pane - it has stored this output as a data frame.

## Returning to the plot
You will recall that the whole reason behind assigning names to this code was to keep it clean and also make plotting more efficient.  
As a result of us assigning the name `beds_ts` to the code segment we can replace:
```{r, eval=FALSE}
ggplot(data = beds_data %>%
         group_by(date)%>%
         summarise(
           mean_beds = mean(beds_av, na.rm = T))
       ) +
  geom...
```

With this:  

```{r, eval=FALSE}
ggplot(data = beds_ts) +
  geom_...(aes(x = ..., y = ...))
```


## Task: Plot the trend in mean available beds.
Slide says: "You may require `ggplot()` plus `+` the following layers:"
```{r, eval = FALSE}
geom_line(aes(..., ...)) +
geom_point(...(..., ...)) +
ylim(..., ...) # To get help with this (or any) function type ?ylim
```

I'm not sure why you'd want `ylim` here (AK). To me, the boundaries look reasonable - maybe they wanted to cut off the earlier entries which were substantially greater? I'm not sure; probably something that would be more apparent in the actual sessions.

```{r}
beds_ts_plot <- ggplot(data = beds_ts) +
  geom_line(aes(x = date, y = mean_beds)) +
  geom_point(aes(x = date, y = mean_beds)) +
  geom_smooth(aes(x = date, y = mean_beds), method = "lm") +
  labs(
    title = "Mean available beds over time",
    x = "Time",
    y = "Mean bed numbers"
  )

beds_ts_plot
```

