---
title: "Intro to R - Wrangling with dplyr"
author: "A.Kirkham"
date: "10/02/2020"
#output: html_document
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
This is a guide to wrangling data uing the dplyr() [dee-ply-r] package.

This guide to wrangling with dplyr() has been built following slides from the NHS-R community at:  
https://github.com/nhs-r-community/intro_r/blob/master/05-workshopv2_dplyr.pdf

The below code and output are generated from following these slides.

All of the resources used here are found on the NHS-R Community Github pages:
https://github.com/nhs-r-community/intro_r


**Now follows a workthrough of the slides.**
 
 Wrangling data means reshaping or transforming data into a format which is easier to work with (for later visualisation, modelling, or computing of statistics).

## Tidyverse

Tidyverse is a group of packages that work in tandem to make processing and assessing data easier.

Tidyverse functions work best with tidy data:  
1. Each variable forms a column  
2. Each observation forms a row  
(Broadly, this means long rather than wide tables)

## The tool: dplyr()
dplyr is a function for data manipulation within the Tidyverse.

Most wrangling can be solved with 5 dplyr verbs:  
**Arrange;**  
**Filter;**  
**Mutate;**  
**Group_By;**  
**Summarise.**  

## Exploring mental health (MH) inpatient capacity
Imagine the situation: We have been asked to conduct an analysis of Mental Health inpatient capacity in England.

The data: KH03 returns (bed numbers and occupancy) by organisation.  
Published by NHSE:
https://www.england.nhs.uk/statistics/statistical-work-areas/bed-availability-and-occupancy/bed-data-overnight/

We will be looking at the changes in the number (and occupancy) of MH beds available in recent years.

**Background**  
Maintaining clinical effectiveness and safety when a ward is fully occupied is a serious challenge for staff.  

Inappropriate out of area placements mean individuals are separated from their social networks for the duration of their care.

**Start the data process**

```{r, message=FALSE}
library(tidyverse)
```

Load the data; **but** use the Import Dataset tool (Environment > Import Dataset > From Text (readr))- this data is not tidy!

Problem with the data: There is metadata above the data we want as the headers ("KH03 returns").  
*Solution is to skip 3 rows*  

But, we also note that the data column is currently set as a character. This needs to be changed to date (%d/%m/%y):

```{r}
library(readr)
beds_data <- read_csv("beds_data.csv", col_types = cols(date = col_date(format = "%d/%m/%Y")), 
                      skip = 3)
View(beds_data)
beds_data
```
(note: `view(beds_data)` loads the dataset and presents it in a new window for interrogation - not shown here.)

It is worth mentioning at this point that the dataset is "real" in that it contains data quality issues galore. These are retained, and we will work with them.

Observations are quarterly.  
*beds_av* == mean number of beds available at midnight over the 3-month period.  
*beds_oc* == as above, occupied.  
Real data - there are problems (NA, etc.)

### Operating dplyr (and tidyverse)
The slides at this point go into greater depth around how a tidyverse syntax is structured. These probably work better in person than in text; therefore I will summarise. (Thankfully this is something that I was already aware of, so it makes it easier to describe).

The slides give a good example, using the cooking of mashed potato:  
We don't just 'get' mashed potato, there are steps involved in the production of it. A recipe, if you will:  
Get potato > peel it > slice it into medium pieces > boil it for 25 minutes > mash it.  

This is exactly how tidyverse syntax works:
We take the data > do something with it > do something else with it > and finally do something else with it again.  

`data_frame THEN #input do_this(rules) THEN #dplyr verb do_this(rules) THEN #dplyr verb do_this(rules) #output (new dataframe)`

However, in tidyverse syntax we don't write THEN we write **%>%** which is known as a **pipe**.

This means that we can combine simple pieces to solve complex issues.

The below questions will help to work through the verbs used in dplyr.

### Q1: Which organisation provided the highest number of MH beds?
1. **arrange** - reorder rows based on a selected variable
```{r}
beds_data %>%
  arrange(desc(beds_av)) #defaults to asc like SQL
```

### Q2: Which organisations provided the highest number of MH beds in Sept. 2018?
We'll use **arrange** as in Q1, but also **filter** in order to select the appropriate data that we need to answer the question.  
Firstly we will take the dataset (beds_data), then we will filter this to only examine where the date is Sept. 2018, and finally arrange it by the availability of beds (mean):

```{r}
beds_data %>%
  filter(date == "2018-09-01") %>%
  arrange(desc(beds_av))
  
```


### Q3: Which 5 organisations had the highest percentage occupancy in Sept. 2018?
As above we will continue to use **arrange** and **filter**, but we do not have all of the necessary information to answer this question. We require another new verb: **mutate**.

**Mutate** serves to create a new variable by mutating other variables. In this instance we do not have the percentage occupancy within the dataset; therefore we will use **mutate** to create it. It is just the same as creating a new value within SQL. To calculate this figure we shall have to divide the occupancy average by the average number of beds. Doing so (using **mutate**) means that we will create a new column (perc_occ) which we can then later use in the same query:

```{r}
beds_data %>%
  mutate(perc_occ = occ_av / beds_av) %>% # Create new variable perc_occ
  filter(date == "2018-09-01") %>%
    arrange(desc(perc_occ)) # Sort by newly created variable
```
Note: Within the above code you will notice that there are multiple routes of using '='. In the instance of **filter** we are assessing for equality (we are only choosing rows where this expression is TRUE). In the instance of **mutate** this is not an assessment of equality but instead we are stating that the new column is equal to the calculation rather than obtaining a TRUE/FALSE outcome only.

### Q4: What was the mean number of beds (across all trusts) for each date of value?
A new verb here: **summarise**. We can use **summarise** to produce summary statistics (like mean), and in doing so collapse many values into a single summary value.  
It has similar syntax to **mutate** but we are not producing any new calculations (mutations), we are going to use data that already exists. We will however still create a new column - again this does not require **mutate**, we can call a new column/variable at any time without **mutate** if the data exists in the required form.

```{r}
beds_data %>%
  summarise(mean_beds = mean(beds_av)) # Problem! There are NA so this would just produce NA
```

The above code clearly does not work! This is because there are NA values within the dataset (the aforementioned data quality issues). Therefore we need to remove these within the syntax. This is wrangling of data. We use the command `na.rm = TRUE` (we could just write T instead of TRUE, it does the same thing) to request that *na* values be *removed*. Essentially "remove NA? True/False".

```{r}
beds_data %>%
  summarise(mean_beds = mean(beds_av, na.rm = TRUE)) # So we remove the NA

```

The above has given a mean overall, a single summary. Now we need to group for the each date element. The question specifically asked for the mean number of beds **for each date**.

The next verb: **group by**. Using the **group by** verb does not directly do anything to the dataframe, but it does change settings behind the scenes. In this case we are wanting to group the data by date, hence we use the syntax `group_by(date)`.

This will result in an output where a value (mean_beds) is shown for each date (a row for each unique value of date).

The slides detail that it is best to `ungroup()` after you have performed the required operation; I don't know why but I have little reason to not believe this...

```{r}
beds_data %>%
  group_by(date) %>%
  summarise(mean_beds = mean(beds_av, na.rm = T)) %>%
  ungroup() # Safest to ungroup after use
```


# Q5: Which 5 organisations have the highest mean % bed occupancy? (over the 5 year period)
For this it is best to break it down into portions:  
1: Create a new variable **mutate** to determine the '% bed occupancy'.  
2: Then, for each of the organisations **group by**.  
3: Then, produce the summary stat (mean) using **summarise**.  
4: Then, use **arrange** to order the output and find the 'highest'.  
Tip: Run the code after each new line to check it returns the output you'd expect.

```{r}
beds_data %>%
  mutate(bed_occ = occ_av / beds_av) %>%
  group_by(org_name) %>%
  summarise(mean_bed_occ = mean(bed_occ, na.rm = T)) %>%
  arrange(desc(mean_bed_occ))  %>%
  ungroup()
```



# Extension: 

### How many columns associated with each observation?
The question posed means that we need an extra column within out summary output. This is not a problem; we can add many extra measures to our summary values. `summarise(number = n())` will count the number of rows associated with each group.  

This can be pushed into the summarise syntax (along with the mean).

```{r}
beds_data %>%
  mutate(bed_occ = occ_av / beds_av) %>%
  group_by(org_name) %>%
  summarise(mean_bed_occ = mean(bed_occ, na.rm = T),
            number = n()) %>%
  arrange(desc(mean_bed_occ)) %>%
  ungroup()
```


## Also a sixth verb: select()

### Select a subset of variables from existing dataset
This **select** verb ensures that we only **select** the data that we are interested in - very much like SELECT in SQL.

For example, the below will return just the fields [org_code] and [org_name]:  
```{r}
beds_data %>%
  select(org_code, org_name)
```

Similarly, we can return all of the fields *except* for [org_code] using the below:  
```{r}
# Select a subset of variables from existing dataset:
beds_data %>%
  select(-org_code) # To remove a column
```

Or, just the first three columns (variables) from the dataset:  
```{r}
beds_data %>%
  select(1:3) # Columns 1 to 3
```

Or, we can return everything but put the [org_name] at the start of the data that is returned:  
```{r}
beds_data %>%
  select(org_name, everything()) # Put org_name at the start of the data frame
```


A couple of pointers that I have observed here:  

* You can combine any number of verbs within the query syntax. For example, in these final points where **select** was used I could have selected only the data that I was interested in, *and then* also performed various other manipulations on the subsequent data. There appears to be no limits.

* Any manipulations (possibly with the exception of `group_by()` which means it is especially important to use `ungroup()`) are kept within the framework of that particular syntax/query. A pipe function appears to only be applicable within the single environment where it is used. For example, if I did a **select** on `beds_data` then that subsequent data frame would only be applicable within that usage environment. If I called `beds_data` within another query/syntax then the **select** manipulation would not be presented. It is almost *non-destructive*, to coin a photography phrase.

# Addendum
## Data frames
In both this portion (dplyr) and other slides (ggplot) we have worked with data frames.

Within a data frame variables are stored in columns, and dataframes are a series of columns bound together by an invisible "data frame" structure.

If we were to remove the data frame structure we can look at the columns in isolation. These isolated columns can be thought of as vectors.

You can create a vector using code like this:
```{r}
c(100, 80, 200) #c stands for "combine"
```